Nested Stochastic Block Models Applied to the Analysis of Single Cell Data

### 2021-04-27 manuscript.tex
Last minute mods

\texttt{schist} is a convenient wrapper to the \texttt{graph-tool} python library, written in python and designed to be used with \texttt{scanpy}. The most prominent function is \texttt{schist.inference.nested\_model()} which takes a \texttt{AnnData} object as input and fits a nested Stochastic Block Model on the \emph{k}NN graph built with \texttt{scanpy} functions (\emph{e.g.} \texttt{scanpy.tools.neighbors()}). When launched with default parameters, \texttt{schist} fits a model which maximises the posterior probability of having a set of cell groups (or blocks) given a graph. \texttt{schist} then annotates cells in the data object with all the groups found at each level of a hierarchy. Given the large size of the NN graph in real-world experiments, it is possible that a single solution represents local minima of the fitting process. In addition, it is possible that multiple solutions are equally acceptable to represent the graph partitioning and a better description is given by the consensus over such solutions \cite{Peixoto_2021_consensus}. To overcome these issues, \texttt{schist} fits multiple instances in parallel and returns the inferred consensus model, alongside the marginal probabilities for each cell to belong to a specific group (\emph{cell marginals}). In addition to the default nested Stochastic Block Model, \texttt{schist} implements the possibility to fit a Bayesian model able to find statistically significant assortative modules \cite{Zhang_Peixoto_2020}, eliminating the need to set a resolution parameter as required in standard community detection by maximisation of modularity.

Once a model has been fitted, \texttt{schist} allows to evaluate the difference in entropy given by assigning every cell to every possible group. This step allows to estimate the strength of association of a cell to a specific group (\emph{cell affinity}). We show that cell affinity may have practical application in common tasks in single cell analysis, such as label transfer.

@@

After preprocessing raw data according to the parameters given in \cite{mereu_2020}, we integrated each dataset with 10XV3 into a unified representation using Harmony \cite{Korsunsky_2019}, and computed the \emph{k}NN graph. In each merged dataset, we retained cell type annotations for 10X cells, while we assigned a "Unknown" label to all cells derived from the other technology (\emph{i.e.} MARS-seq or Quartz-seq2). We then calculated the cell affinity matrix, that is we computed the difference in entropy that can be observed by assigning each cell to each annotation cluster, this being either one of the original cell types or "Unknown". Once the matrix has been computed, each cell from the query data is assigned to the group with the highest likelihood. The rationale behind this approach is that if cells belong to the same annotation group, then more information is required to describe the graph if they were annotated as different cell types; hence, cells from the query datasets should retain their "Unknown" label if and only if there is not enough evidence to associate them to another group. We compared the accuracy of the  outcome to \emph{k}NN classification, given by the closest entry in the \emph{k}NN graph, and to \texttt{ingest}, a tool included in \texttt{scanpy}  based on \emph{k}NN classification of UMAP embeddings. Analysis of a well defined dataset, such as Quartz-seq2, reveals that the three approaches are equally good in classifying unknown cells (Fig. \ref{Figure_Label_Transfer}, central column), with accuracies ranging from .870 to .927. When data are noisy, instead, \emph{k}NN-based methods show low accuracy and a tendency to assign the most represented cell group (HEK293T) to the unlabelled cells. This misannotation is particular evident for \texttt{ingest}, in which only CD4 T cells and HEK cells are transferred, resulting in the lowest accuracy (0.243). Conversely, \texttt{schist} is able to assign correct labels with higher accuracy (0.641). Moreover, \emph{k}NN methods assign a label to each cell, whereas \texttt{schist} does not relabel cells if there are no sufficient evidence (\emph{e.g.} the "Unknown" state is the most likely). Interestingly, we found that for the largest part of cells without assigned label, the second choice by affinity ranking was indeed the appropriate one (Fig. \ref{hist_label_second}).

>>>>

\texttt{schist} is a convenient wrapper to the \texttt{graph-tool} python library, written in python and designed to be used with \texttt{scanpy}. The most prominent function is \texttt{schist.inference.nested\_model()} which takes a \texttt{AnnData} object as input and fits a nested Stochastic Block Model on the \emph{k}NN graph built with \texttt{scanpy} functions (\emph{e.g.} \texttt{scanpy.tools.neighbors()}). When launched with default parameters, \texttt{schist} fits a model which maximises the posterior probability of having a set of cell groups (or blocks) given a graph. \texttt{schist} then annotates cells in the data object with all the groups found at each level of a hierarchy. Given the large size of the NN graph in real-world experiments, it is possible that a single solution represents local minima of the fitting process. In addition, it is possible that multiple solutions are equally acceptable to represent the graph partitioning and a better description is given by the consensus over such solutions \cite{Peixoto_2021_consensus}. To overcome these issues, \texttt{schist} fits multiple instances in parallel and returns the inferred consensus model, alongside the marginal probabilities for each cell to belong to a specific group (\emph{cell marginals}). 
Moreover, the Stochastic Block Model has no constraints on what type of modular structure is fitted, meaning that groups are not necessarily identified only by assortativity (\emph{i.e.} cells are mostly connected within the same group). When assortativity is thought to be the dominant pattern another model (the Planted Partition Block Model, PPBM \cite{Zhang_Peixoto_2020}), also implemented in \texttt{schist}, is better suited to find statistically significant assortative communities, also eliminating the need to set a resolution parameter as required in standard community detection by maximisation of modularity.

%Once a model has been fitted, \texttt{schist} allows to evaluate the difference in entropy given by assigning every cell to every possible group. This step allows to estimate the strength of association of a cell to a specific group (\emph{cell affinity}). We show that cell affinity may have practical application in common tasks in single cell analysis, such as label transfer.

@@

After preprocessing raw data according to the parameters given in \cite{mereu_2020}, we integrated each dataset with 10XV3 into a unified representation using Harmony \cite{Korsunsky_2019}, and computed the \emph{k}NN graph. In each merged dataset, we retained cell type annotations for 10X cells, while we assigned a "Unknown" label to all cells derived from the other technology (\emph{i.e.} MARS-seq or Quartz-seq2). We then calculated the \emph{cell affinity} matrix, that is we computed the difference in entropy that can be observed by assigning each cell to each annotation cluster, this being either one of the original cell types or "Unknown". Once the matrix has been computed, each cell from the query data is assigned to the group with the highest likelihood. The rationale behind this approach is that if cells belong to the same annotation group, then more information is required to describe the graph if they were annotated as different cell types; hence, cells from the query datasets should retain their "Unknown" label if and only if there is not enough evidence to associate them to another group. We compared the accuracy of the  outcome to \emph{k}NN classification, given by the closest entry in the \emph{k}NN graph, and to \texttt{ingest}, a tool included in \texttt{scanpy}  based on \emph{k}NN classification of UMAP embeddings. Analysis of a well defined dataset, such as Quartz-seq2, reveals that the three approaches are equally good in classifying unknown cells (Fig. \ref{Figure_Label_Transfer}, central column), with accuracies ranging from .870 to .927. When data are noisy, instead, \emph{k}NN-based methods show low accuracy and a tendency to assign the most represented cell group (HEK293T) to the unlabelled cells. This misannotation is particular evident for \texttt{ingest}, in which only CD4 T cells and HEK cells are transferred, resulting in the lowest accuracy (0.243). Conversely, \texttt{schist} is able to assign correct labels with higher accuracy (0.641). Moreover, \emph{k}NN methods assign a label to each cell, whereas \texttt{schist} does not relabel cells if there are no sufficient evidence (\emph{e.g.} the "Unknown" state is the most likely). Interestingly, we found that for the largest part of cells without assigned label, the second choice by affinity ranking was indeed the appropriate one (Fig. \ref{hist_label_second}).
